Lenses with a Difference

### 2018-06-29 papers/DifferenceLenses/main.tex

Each one of these solutions has significant shortcomings. Firstly, while matching lenses successfully address the issue of alignment, they do not address the general issue of correct update propagation. Secondly, edit lenses and delta lenses all assume that the lens propagates a specific set of updates explicitly, but this is unrealistic since since in general users do not update data using a small predetermined langauge of edits; instead users edit the data in any way, for example using text editors or IDEs, and present the updatad data to the bx wholesale. Finally, least change lenses are in general computationally intractable \cite{buneman2002propagation} and also lack a notion of composition---that is, the composition of two deterministic least change lenses is not always a least change lens\cite{macedo2013composing}.

@@

The functions $\delta_S$ and $\delta_V$ will depend only on $S$ and $V$, and will be designed to measure how similar two sources (or views) are to each other. More specifically, given a regular expression $R$ and strings $s, s' \in \mathcal{L}(R)$, we will define a number $\delta_R(s, s')$ inductively on $R$, with $0 \leq \delta_R(s, s') \leq 1$, and a high value of $\delta_R(s, s')$ indicates that $s$ and $s'$ are similar---in particular, if $s = s'$, then $\delta_R(s, s') = 1$, and if $s \neq s'$ and either $s = \varepsilon$ or $s' = \varepsilon$, then $\delta_R(s, s') = 0$. We will also expand our definition of the regular expression $R^*$, having three different variants $R^*_{POS}$, $R^*_{NC}$ and $R^*_{MED}$, where the differences are computed using positional, non-crossing and min-edit distance alignment strategies respectively, so that for all $s, s' \in \mathcal{L}(R^*)$, $\delta_{R^*_{MED}}(s, s') \geq \delta_{R^*_{NC}}(s, s') \geq \delta_{R^*_{MED}}(s, s')$.
\subsection{Difference Functions on Regular Languages}
\begin{definition}
Given a finite alphabet $\Sigma$, a {\em difference regular expression} is any expression generated by the following grammar,
$$R := s \; | \; R_1 \cdot R_2 \; | \; R_1 \cup R_2 \; | \; R^*_{POS} \; | \; R^*_{NC} \; | \; R^*_{MED}$$
where $s$ ranges over $\Sigma^*$ and $R$ ranges over regular expressions. The languages of $R^*_{POS}$, $R^*_{NC}$ and $R^*_{MED}$ are all the same, namely $\mathcal{L}(R^*)$.
\end{definition}
\begin{definition}
Given an unambiguous difference regular expression $R$, define $\delta_R : \mathcal{L}(R) \times \mathcal{L}(R) \longrightarrow \mathbb{R}^{\geq 0}$ inductively as follows:
\begin{align*}
\delta_s(t, t') &= 1\\
\delta_{R_1 \cdot R_2}(s_1 \cdot s_2, s'_1 \cdot s'_2) &= \frac{1}{2}(\delta_{R_1}(s_1, s'_1) + \delta_{R_2}(s_2, s'_2))\\
\delta_{R_1 \cup R_2}(s, s') &= 
\begin{cases}
\delta_{R_1}(s, s') & \text{if }s, s' \in \mathcal{L}(R_1) \\
\delta_{R_2}(s, s') & \text{if }s, s' \in \mathcal{L}(R_2)\\
0 & \text{otherwise }
\end{cases}\\
\delta_{R^*_{POS}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
\begin{cases}
1 & \text{if } m = 0, n = 0\\ 
0 & \text{if } m = 0, n \neq 0\\
0 & \text{if } m \neq 0, n = 0\\\max\limits_{\{\varphi : [m] \longrightarrow [n] \; | \; \varphi(i) = i\}}\frac{1}{n}\sum\limits_{i=1}^{m}\ \delta_R(s_i, t_{\varphi(i)})& \text{if } m \leq n\\
\max\limits_{\{\varphi : [n] \longrightarrow [m] \; | \; \varphi(i) = i\}}\frac{1}{m}\sum\limits_{i=1}^{m}\ \delta_R(t_i, s_{\varphi(i)})& \text{otherwise}
\end{cases}\\
\delta_{R^*_{NC}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
\begin{cases}
1 & \text{if } m = 0, n = 0\\ 
0 & \text{if } m = 0, n \neq 0\\
0 & \text{if } m \neq 0, n = 0\\
\max\limits_{\{\varphi : [m] \longrightarrow [n] \; | \; \varphi \text{ is strictly increasing}\}}\frac{1}{n}\sum\limits_{i=1}^{m}\ \delta_R(s_i, t_{\varphi(i)})& \text{if } m \leq n\\
\max\limits_{\{\varphi : [n] \longrightarrow [m] \; | \; \varphi \text{ is strictly increasing}\}}\frac{1}{m}\sum\limits_{i=1}^{m}\ \delta_R(t_i, s_{\varphi(i)})& \text{otherwise}
\end{cases}\\
\delta_{R^*_{MED}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
\begin{cases}
1 & \text{if } m = 0, n = 0\\ 
0 & \text{if } m = 0, n \neq 0\\
0 & \text{if } m \neq 0, n = 0\\\max\limits_{\{\varphi : [m] \longrightarrow [n] \; | \; \varphi \text{ is injective}\}}\frac{1}{n}\sum\limits_{i=1}^{m}\ \delta_R(s_i, t_{\varphi(i)})& \text{if } m \leq n\\
\max\limits_{\{\varphi : [n] \longrightarrow [m] \; | \; \varphi \text{ is injective}\}}\frac{1}{m}\sum\limits_{i=1}^{m}\ \delta_R(t_i, s_{\varphi(i)})& \text{otherwise}
\end{cases}
\end{align*}
\end{definition}
\begin{fact}
For any regular expression $R$ and $s, s' \in \mathcal{L}(R)$
\begin{enumerate}
\item
$0 \leq \delta_R(s, s') \leq 1$.
\item
$\delta_R(s, s) = 1$.
\item
For all $s, s' \in \mathcal{L}(R^*)$, $\delta_{R^*_{MED}}(s, s') \geq \delta_{R^*_{NC}}(s, s') \geq \delta_{R^*_{MED}}(s, s')$.
\item
$\delta_R(s, s') = \delta_R(s', s)$.
\end{enumerate}

@@

\begin{claim}
If $\ell_1 : S_1 \Leftrightarrow V_1$ and $\ell_2 : S_2 \Leftrightarrow V_2$ are difference lenses, with $\mathcal{L}(S_1) \cap \mathcal{L}(S_2) = \varnothing$ and $\mathcal{L}(V_1) \cap \mathcal{L}(V_2) = \varnothing$ then $\ell_1 \cup \ell_2 :
S_1 \cup S_2 \Leftrightarrow V_1 \cup V_2$ is also a difference lens.
\end{claim}
\begin{proof}
If the source and the view are in $S_i$ and $V_i$ respectively for $i \in \{1,2\}$, then the result holds from the fact that $\ell_1$ and $\ell_2$ are difference lenses, otherwise the difference on both the source and view side are equal to 0, from which the result follows.
\end{proof}
\subsubsection{Iteration}
Let $\ell : S \Leftrightarrow V$ is a difference lens with $S^{!*}$ and $V^{!*}$. Let $s_1 \cdot \ldots \cdot s_m$ be a source, and $v_1 \cdot \ldots \cdot v_n$ a view. If $n \leq m$ and $\varphi : [n] \longrightarrow [m]$, define $ v \xhookrightarrow{\ell, \varphi} s = s'_1 \cdot \ldots \cdot s'_n$ by 
$$s'_i = \begin{cases}
\ell.\pput \; s_{\varphi(i)} \; v_i & \text{if } i \in Dom(\varphi)\\
\ell.\create \; v_i & \text{if } i \not \in Dom(\varphi)
\end{cases}$$
If $m < n$ and $\varphi : [m] \longrightarrow [n]$, define $ v \xhookrightarrow{\ell, \varphi} s = s'_1 \cdot \ldots \cdot s'_n$ by 
$$s'_i = \begin{cases}
\ell.\pput \; s_{\varphi^{-1}(i)} \; v_i & \text{if } i \in \varphi([m])\\
\ell.\create \; v_i & \text{if } i \not \in \varphi([m])
\end{cases}$$

Using the notation $len(v)$ and $len(s)$ to mean $len(v) = n$ if $v = v_1 \cdot \ldots \cdot v_n$ and $len(s) = m$ if $s = s_1 \cdot \ldots \cdot s_m$, define the lenses $\ell^*_{POS} : S^*_{POS} \Leftrightarrow V^*_{POS}$, $\ell^*_{NC} : S^*_{NC} \Leftrightarrow V^*_{NC}$ and $\ell^*_{MED} : S^*_{MED} \Leftrightarrow V^*_{MED}$ by:
\begin{align*}
\ell^*_{POS}.\get \; (s_1 \cdot  \ldots \cdot s_n) &= (\ell.\get \; s_1)
\cdot \ldots \cdot (\ell.\get \; s_n)\\
\ell^*_{POS}.\pput \; s \; v
&= v \xhookrightarrow{\psi} s\\
\text{where } \psi &= \begin{cases}
\argmax \limits_{\{\varphi : [len(v)] \longrightarrow [len(s)] \; | \; \varphi(i) = i\}} \delta_{S^*_{NC}}(v \xhookrightarrow{\ell, \varphi} s, s) & \text{if } len(v) \leq len(s)\\
\argmax \limits_{\{\varphi : [len(s)] \longrightarrow [len(v)] \; | \; \varphi(i) = i\}} \delta_{S^*_{NC}}(v \xhookrightarrow{\ell, \varphi} s, s) & \text{if } len(s) < len(v)
\end{cases}\\
\ell^*_{POS}.\create \; (v_1, \ldots, v_n) &= (\ell.\create \;
v_1) \cdot \ldots \cdot (\ell.\create \; v_n)
\end{align*}
\begin{align*}
\ell^*_{NC}.\get \; (s_1 \cdot  \ldots \cdot s_n) &= (\ell.\get \; s_1)
\cdot \ldots \cdot (\ell.\get \; s_n)\\
\ell^*_{NC}.\pput \; s \; v
&= v \xhookrightarrow{\psi} s\\
\text{where } \psi &= \begin{cases}
\argmax \limits_{\{\varphi : [len(v)] \longrightarrow [len(s)] \; | \; \varphi \text{ is str. increasing}\}} \delta_{S^*_{NC}}(v \xhookrightarrow{\ell, \varphi} s, s) & \text{if } len(v) \leq len(s)\\
\argmax \limits_{\{\varphi : [len(s)] \longrightarrow [len(v)] \; | \; \varphi \text{ is str. increasing}\}} \delta_{S^*_{NC}}(v \xhookrightarrow{\ell, \varphi} s, s) & \text{if } len(s) < len(v)
\end{cases}\\
\ell^*_{NC}.\create \; (v_1, \ldots, v_n) &= (\ell.\create \;
v_1) \cdot \ldots \cdot (\ell.\create \; v_n)
\end{align*}
\begin{align*}
\ell^*_{MED}.\get \; (s_1 \cdot  \ldots \cdot s_n) &= (\ell.\get \; s_1)
\cdot \ldots \cdot (\ell.\get \; s_n)\\
\ell^*_{MED}.\pput \; s \; v
&= v \xhookrightarrow{\ell, \psi} s\\
\text{where } \psi &= \begin{cases}
\argmax \limits_{\{\varphi : [len(v)] \longrightarrow [len(s)] \; | \; \varphi \text{ is injective}\}} \delta_{S^*_{NC}}(v \xhookrightarrow{\ell, \varphi} s, s) & \text{if } len(v) \leq len(s)\\
\argmax \limits_{\{\varphi : [len(s)] \longrightarrow [len(v)] \; | \; \varphi \text{ is injective}\}} \delta_{S^*_{NC}}(v \xhookrightarrow{\ell, \varphi} s, s) & \text{if } len(s) < len(v)
\end{cases}\\
\ell^*_{MED}.\create \; (v_1, \ldots, v_n) &= (\ell.\create \;
v_1) \cdot \ldots \cdot (\ell.\create \; v_n)
\end{align*}
For the non-crossing and min-edit distance lens, there may be multiple alignments $\psi$ that maximize the respective differences, in which case the lens chooses the lexicographically smallest of those alignments.

\begin{claim}
If $\ell : S \Leftrightarrow V$ is a difference lens with $S^{!*}$ and $V^{!*}$, then $\ell^*_{POS}$, $\ell^*_{NC}$ and $\ell^*_{MED}$ are also difference lenses.
\end{claim}
\begin{proof}
We will show that $\ell^*_{MED}$ is a difference lens; the proof for each of the other lenses is similar.

We need to show that
$$\delta_{S^*_{MED}}(\ell^*_{MED}.\pput \; (s_1 \cdot \ldots \cdot s_m) \; (v_1 \cdot \ldots \cdot v_n), s_1 \cdot \ldots \cdot s_m) \geq \delta_{V^*_{MED}}(v_1 \cdot \ldots \cdot v_n, (\ell.\get \; s_1) \cdot \ldots (\ell.\get \; s_m))$$
Assume wthout loss of generality that $n \leq m$. An optimal min-edit distance alignment $\varphi : [n] \longrightarrow [m]$ is used to compute the difference on the view side. Since $\ell$ is a difference lens, the following equation holds,
$$\delta_S(\ell.\pput \; s_{\varphi(i)} \; v_i, s_{\varphi(i)}) \geq \delta_V(v_i, \ell.\get \; s_{\varphi(i)})$$
from which it follows that
$$\delta_{S^*_{MED}}(v_1 \cdot \ldots v_n \xhookrightarrow{\ell, \varphi} s_1 \cdot \ldots \cdot s_m, s_1 \cdot \ldots \cdot s_m) \geq \delta_{V^*_{NC}}(v_1 \cdot \ldots \cdot v_n, (\ell.\get \; s_1) \cdot \ldots (\ell.\get \; s_m))$$
Since the \pput function of $\ell^*_{MED}$ uses an alignment $\psi$ that maximizes the min-edit distance over all injective alignments mapping $v$ into $s$, then
$$\delta_{S^*_{MED}}(v_1 \cdot \ldots v_n \xhookrightarrow{\ell, \psi} s_1 \cdot \ldots \cdot s_m, s_1 \cdot \ldots \cdot s_m) \geq \delta_{S^*_{MED}}(v_1 \cdot \ldots v_n \xhookrightarrow{\varphi} s_1 \cdot \ldots \cdot s_m, s_1 \cdot \ldots \cdot s_m)$$

It thus follows that,
$$\delta_{S^*_{MED}}(v_1 \cdot \ldots v_n \xhookrightarrow{\ell, \psi} s_1 \cdot \ldots \cdot s_m, s_1 \cdot \ldots \cdot s_m) \geq \delta_{V^*_{NC}}(v_1 \cdot \ldots \cdot v_n, (\ell.\get \; s_1) \cdot \ldots (\ell.\get \; s_m))$$
hence,
$$\delta_{S^*_{MED}}(\ell^*_{MED}.\pput \; (s_1 \cdot \ldots \cdot s_m) \; (v_1 \cdot \ldots \cdot v_n), s_1 \cdot \ldots \cdot s_m) \geq \delta_{V^*_{MED}}(v_1 \cdot \ldots \cdot v_n, (\ell.\get \; s_1) \cdot \ldots (\ell.\get \; s_m))$$
which is what we wanted to show.
\end{proof}
\subsubsection{Subtyping}
Consider a positional iteration lens $\ell^*_{POS} : S^*_{POS} \Leftrightarrow V^*_{POS}$. Since $\delta_{S^*_{NC}}(s, s') \geq \delta_{S^*_{POS}}(s, s')$ for all $s, s' \in \mathcal{L}(R)$, it follows $\ell^*_{POS} : S^*_{NC} \Leftrightarrow V^*_{POS}$ is also a valid typing judgement. This observation suggests a {\em subtyping} relation on unambiguous difference regular expressions.

@@

For any unambiguous regular expression $R$, let $List(R)$ be the set of all lists with elements in $\mathcal{L}(R)$, and observe that we can define three difference functions $\delta^{List(R)}_{POS}$ $\delta^{List(R)}_{NC}$ and $\delta^{List(R)}_{MED}$ on $List(R)$ in the same way that we define $\delta_{R^*_{POS}}$, $\delta_{R^*_{NC}}$ and $\delta_{R^*_{MED}}$. For example, define
$$
\delta^{List(R)}_{MED}([s_1, \ldots, s_m], [t_1, \ldots, t_n]) = 
\begin{cases}
1 & \text{if } m = 0, n = 0\\ 
0 & \text{if } m = 0, n \neq 0\\
0 & \text{if } m \neq 0, n = 0\\\max\limits_{\{\varphi : [m] \longrightarrow [n] \; | \; \varphi \text{ is injective}\}}\frac{1}{n}\sum\limits_{i=1}^{m}\ \delta_R(s_i, t_{\varphi(i)})& \text{if } m \leq n\\
\max\limits_{\{\varphi : [n] \longrightarrow [m] \; | \; \varphi \text{ is injective}\}}\frac{1}{m}\sum\limits_{i=1}^{m}\ \delta_R(t_i, s_{\varphi(i)})& \text{otherwise}
\end{cases}
$$

>>>>

Each one of these solutions has significant shortcomings. While matching lenses successfully address the issue of alignment, they do not address the general issue of correct update propagation. Edit lenses and delta lenses all assume that the lens propagates a specific set of updates explicitly, which is unrealistic since since in general users do not update data using a small predetermined langauge of edits; instead they edit the data in any way, for example using text editors or IDEs, and present the updatad data to the bx wholesale. Finally, least change lenses are in general computationally intractable \cite{buneman2002propagation} and also lack a notion of composition---that is, the composition of two deterministic least change lenses is not always a least change lens\cite{macedo2013composing}.

@@

The functions $\delta_S$ and $\delta_V$ will depend only on $S$ and $V$, and will be designed to measure how similar two sources (or views) are to each other. More specifically, given a regular expression $R$ and strings $s, s' \in \mathcal{L}(R)$, we will define a number $\delta_R(s, s')$ inductively on $R$, with $0 \leq \delta_R(s, s') \leq 1$, and a high value of $\delta_R(s, s')$ indicates that $s$ and $s'$ are similar---in particular, if $s = s'$, then $\delta_R(s, s') = 1$, and if $s \neq s'$ and either $s = \varepsilon$ or $s' = \varepsilon$, then $\delta_R(s, s') = 0$. We will also expand our definition of the regular expression $R^*$, having three different variants $R^*_{POS}$, $R^*_{NC}$ and $R^*_{MED}$, where the differences are computed using positional, non-crossing and min-edit distance alignment strategies respectively, so that for all $s, s' \in \mathcal{L}(R^*)$, $\delta_{R^*_{MED}}(s, s') \geq \delta_{R^*_{NC}}(s, s') \geq \delta_{R^*_{MED}}(s, s')$.
\subsection{Difference Functions on Regular Languages}
\begin{definition}
Given a finite alphabet $\Sigma$, a {\em difference regular expression} is any expression generated by the following grammar,
$$R := s \; | \; R_1 \cdot R_2 \; | \; R_1 \cup R_2 \; | \; R^*_{POS} \; | \; R^*_{NC} \; | \; R^*_{MED}$$
where $s$ ranges over $\Sigma^*$ and $R$ ranges over regular expressions. The languages of $R^*_{POS}$, $R^*_{NC}$ and $R^*_{MED}$ are all the same, namely $\mathcal{L}(R^*)$.
\end{definition}
\begin{definition}
Given an unambiguous difference regular expression $R$, define $\delta_R : \mathcal{L}(R) \times \mathcal{L}(R) \longrightarrow \mathbb{R}^{\geq 0}$ inductively as follows:
\begin{align*}
\delta_s(t, t') &= 1\\
\delta_{R_1 \cdot R_2}(s_1 \cdot s_2, s'_1 \cdot s'_2) &= \frac{1}{2}(\delta_{R_1}(s_1, s'_1) + \delta_{R_2}(s_2, s'_2))\\
\delta_{R_1 \cup R_2}(s, s') &= 
\begin{cases}
\delta_{R_1}(s, s') & \text{if }s, s' \in \mathcal{L}(R_1) \\
\delta_{R_2}(s, s') & \text{if }s, s' \in \mathcal{L}(R_2)\\
0 & \text{otherwise }
\end{cases}\\
\delta_{R^*_{POS}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
\begin{cases}
1 & \text{if } m = 0, n = 0\\ 
0 & \text{if } m = 0, n \neq 0\\
0 & \text{if } m \neq 0, n = 0\\\max\limits_{\{\varphi : [m] \rightharpoonup [n] \; | \; \varphi(i) = i\}}\sum\limits_{i \in Dom(\varphi)}\ \frac{\delta_R(s_i, t_{\varphi(i)})}{\max(m,n)}& \text{otherwise}
\end{cases}\\
\delta_{R^*_{NC}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
\begin{cases}
1 & \text{if } m = 0, n = 0\\ 
0 & \text{if } m = 0, n \neq 0\\
0 & \text{if } m \neq 0, n = 0\\
\max\limits_{\{\varphi : [m] \rightharpoonup [n] \; | \; \varphi \text{ is str. increasing}\}}\sum\limits_{i \in Dom(\varphi)}\ \frac{\delta_R(s_i, t_{\varphi(i)})}{\max(m,n)}& \text{otherwise}
\end{cases}\\
\delta_{R^*_{MED}}(s_1 \cdot \ldots \cdot s_m, t_1 \cdot \ldots \cdot t_n) &= 
\begin{cases}
1 & \text{if } m = 0, n = 0\\ 
0 & \text{if } m = 0, n \neq 0\\
0 & \text{if } m \neq 0, n = 0\\\max\limits_{\{\varphi : [m] \rightharpoonup [n] \; | \; \varphi \text{ is injective}\}}\sum\limits_{i \in Dom(\varphi)}\ \frac{\delta_R(s_i, t_{\varphi(i)})}{\max(m,n)}& \text{otherwise }
\end{cases}
\end{align*}
\end{definition}
\begin{fact}
For any regular expression $R$ and $s, s' \in \mathcal{L}(R)$
\begin{enumerate}
\item
$0 \leq \delta_R(s, s') \leq 1$.
\item
$\delta_R(s, s) = 1$.
\item
$\delta_R(s, s') = \delta_R(s', s)$.
\item
For all $s, s' \in \mathcal{L}(R^*)$, $\delta_{R^*_{MED}}(s, s') \geq \delta_{R^*_{NC}}(s, s') \geq \delta_{R^*_{MED}}(s, s')$.
\end{enumerate}

@@

\begin{claim}
If $\ell_1 : S_1 \Leftrightarrow V_1$ and $\ell_2 : S_2 \Leftrightarrow V_2$ are difference lenses, with $\mathcal{L}(S_1) \cap \mathcal{L}(S_2) = \varnothing$ and $\mathcal{L}(V_1) \cap \mathcal{L}(V_2) = \varnothing$ then $\ell_1 \cup \ell_2 :
S_1 \cup S_2 \Leftrightarrow V_1 \cup V_2$ is also a difference lens.
\end{claim}
\begin{proof}
If the source and the view are in $S_i$ and $V_i$ respectively for $i \in \{1,2\}$, then the result holds from the fact that $\ell_1$ and $\ell_2$ are difference lenses, otherwise the difference on both the source and view side are equal to 0, from which the result follows.
\end{proof}
\subsubsection{Iteration}
Let $\ell : S \Leftrightarrow V$ is a difference lens with $S^{!*}$ and $V^{!*}$. Let $s_1 \cdot \ldots \cdot s_m$ be a source, and $v_1 \cdot \ldots \cdot v_n$ a view. Let $\varphi : [n] \rightharpoonup [m]$ and define $ v \xhookrightarrow{\ell, \varphi} s = s'_1 \cdot \ldots \cdot s'_n$ by 
$$s'_i = \begin{cases}
\ell.\pput \; s_{\varphi(i)} \; v_i & \text{if } i \in Dom(\varphi)\\
\ell.\create \; v_i & \text{if } i \not \in Dom(\varphi)
\end{cases}$$

Using the notation $len(v)$ and $len(s)$ to mean $len(v) = n$ if $v = v_1 \cdot \ldots \cdot v_n$ and $len(s) = m$ if $s = s_1 \cdot \ldots \cdot s_m$, define the lenses $\ell^*_{POS} : S^*_{POS} \Leftrightarrow V^*_{POS}$, $\ell^*_{NC} : S^*_{NC} \Leftrightarrow V^*_{NC}$ and $\ell^*_{MED} : S^*_{MED} \Leftrightarrow V^*_{MED}$ by:
\begin{align*}
\ell^*_{POS}.\get \; (s_1 \cdot  \ldots \cdot s_n) &= (\ell.\get \; s_1)
\cdot \ldots \cdot (\ell.\get \; s_n)\\
\ell^*_{POS}.\pput \; s \; v
&= v \xhookrightarrow{\psi} s\\
\text{where } \psi &= 
\argmax \limits_{\{\varphi : [len(v)] \rightharpoonup [len(s)] \; | \; \varphi(i) = i\}} \delta_{S^*_{POS}}(v \xhookrightarrow{\ell, \varphi} s, s)\\
\ell^*_{POS}.\create \; (v_1, \ldots, v_n) &= (\ell.\create \;
v_1) \cdot \ldots \cdot (\ell.\create \; v_n)
\end{align*}
\begin{align*}
\ell^*_{NC}.\get \; (s_1 \cdot  \ldots \cdot s_n) &= (\ell.\get \; s_1)
\cdot \ldots \cdot (\ell.\get \; s_n)\\
\ell^*_{NC}.\pput \; s \; v
&= v \xhookrightarrow{\psi} s\\
\text{where } \psi &= 
\argmax \limits_{\{\varphi : [len(v)] \rightharpoonup [len(s)] \; | \; \varphi \text{ is str. increasing}\}} \delta_{S^*_{NC}}(v \xhookrightarrow{\ell, \varphi} s, s) \\
\ell^*_{NC}.\create \; (v_1, \ldots, v_n) &= (\ell.\create \;
v_1) \cdot \ldots \cdot (\ell.\create \; v_n)
\end{align*}
\begin{align*}
\ell^*_{MED}.\get \; (s_1 \cdot  \ldots \cdot s_n) &= (\ell.\get \; s_1)
\cdot \ldots \cdot (\ell.\get \; s_n)\\
\ell^*_{MED}.\pput \; s \; v
&= v \xhookrightarrow{\ell, \psi} s\\
\text{where } \psi &= 
\argmax \limits_{\{\varphi : [len(v)] \rightharpoonup [len(s)] \; | \; \varphi \text{ is injective}\}} \delta_{S^*_{MED}}(v \xhookrightarrow{\ell, \varphi} s, s)\\
\ell^*_{MED}.\create \; (v_1, \ldots, v_n) &= (\ell.\create \;
v_1) \cdot \ldots \cdot (\ell.\create \; v_n)
\end{align*}
\begin{claim}
If $\ell : S \Leftrightarrow V$ is a difference lens with $S^{!*}$ and $V^{!*}$, then $\ell^*_{POS}$, $\ell^*_{NC}$ and $\ell^*_{MED}$ are also difference lenses.
\end{claim}
\begin{proof}
We will show that $\ell^*_{MED}$ is a difference lens; the proof for each of the other lenses is similar.

We need to show that
$$\delta_{S^*_{MED}}(\ell^*_{MED}.\pput \; (s_1 \cdot \ldots \cdot s_m) \; (v_1 \cdot \ldots \cdot v_n), s_1 \cdot \ldots \cdot s_m) \geq \delta_{V^*_{MED}}(v_1 \cdot \ldots \cdot v_n, (\ell.\get \; s_1) \cdot \ldots (\ell.\get \; s_m))$$
An optimal min-edit distance alignment $\varphi : [n] \rightharpoonup [m]$ is used to compute the difference on the view side. Since $\ell$ is a difference lens, the following equation holds,
$$\delta_S(\ell.\pput \; s_{\varphi(i)} \; v_i, s_{\varphi(i)}) \geq \delta_V(v_i, \ell.\get \; s_{\varphi(i)})$$
from which it follows that
$$\delta_{S^*_{MED}}(v_1 \cdot \ldots v_n \xhookrightarrow{\ell, \varphi} s_1 \cdot \ldots \cdot s_m, s_1 \cdot \ldots \cdot s_m) \geq \delta_{V^*_{MED}}(v_1 \cdot \ldots \cdot v_n, (\ell.\get \; s_1) \cdot \ldots (\ell.\get \; s_m))$$
Since the \pput function of $\ell^*_{MED}$ uses an alignment $\psi$ that maximizes the min-edit distance over all injective alignments mapping $v$ into $s$, then
$$\delta_{S^*_{MED}}(v_1 \cdot \ldots v_n \xhookrightarrow{\ell, \psi} s_1 \cdot \ldots \cdot s_m, s_1 \cdot \ldots \cdot s_m) \geq \delta_{S^*_{MED}}(v_1 \cdot \ldots v_n \xhookrightarrow{\varphi} s_1 \cdot \ldots \cdot s_m, s_1 \cdot \ldots \cdot s_m)$$

It thus follows that,
$$\delta_{S^*_{MED}}(v_1 \cdot \ldots v_n \xhookrightarrow{\ell, \psi} s_1 \cdot \ldots \cdot s_m, s_1 \cdot \ldots \cdot s_m) \geq \delta_{V^*_{MED}}(v_1 \cdot \ldots \cdot v_n, (\ell.\get \; s_1) \cdot \ldots (\ell.\get \; s_m))$$
hence,
$$\delta_{S^*_{MED}}(\ell^*_{MED}.\pput \; (s_1 \cdot \ldots \cdot s_m) \; (v_1 \cdot \ldots \cdot v_n), s_1 \cdot \ldots \cdot s_m) \geq \delta_{V^*_{MED}}(v_1 \cdot \ldots \cdot v_n, (\ell.\get \; s_1) \cdot \ldots (\ell.\get \; s_m))$$
which is what we wanted to show.
\end{proof}
\subsubsection{Subtyping}
Consider a positional iteration lens $\ell^*_{POS} : S^*_{POS} \Leftrightarrow V^*_{POS}$. Since $\delta_{S^*_{NC}}(s, s') \geq \delta_{S^*_{POS}}(s, s')$ for all $s, s' \in \mathcal{L}(R)$, it follows $\ell^*_{POS} : S^*_{NC} \Leftrightarrow V^*_{POS}$ is also a valid typing judgement. This observation suggests a {\em subtyping} relation on unambiguous difference regular expressions.

@@

For any unambiguous regular expression $R$, let $List(R)$ be the set of all lists with elements in $\mathcal{L}(R)$, and observe that we can define three difference functions $\delta^{List(R)}_{POS}$ $\delta^{List(R)}_{NC}$ and $\delta^{List(R)}_{MED}$ on $List(R)$ in the same way that we define $\delta_{R^*_{POS}}$, $\delta_{R^*_{NC}}$ and $\delta_{R^*_{MED}}$. For example, define
$$
\delta^{List(R)}_{MED}([s_1, \ldots, s_m], [t_1, \ldots, t_n]) = 
\begin{cases}
1 & \text{if } m = 0, n = 0\\ 
0 & \text{if } m = 0, n \neq 0\\
0 & \text{if } m \neq 0, n = 0\\\max\limits_{\{\varphi : [m] \rightharpoonup [n] \; | \; \varphi \text{ is injective}\}}\sum\limits_{i=1}^{m}\ \frac{\delta_R(s_i, t_{\varphi(i)})}{\max(m,n)}& \text{otherwise}
\end{cases}
$$
